Classification and Regression Trees and Beyond

### 2020-06-30 pkg/partykit/inst/ULGcourse-2020/slides_tree.Rnw
almost finished motivation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3021 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c4d4400c491c5d2ee06160f4ef354a5d2b4f8bfe

{\bf Idea:} \\
\smallskip
\quad Separate data into smaller and more homogeneous subgroups \\
\quad based on information provided by covariates.
%in order to obtain better fitting \\
%\quad models in the resulting subsets.

@@

@@

{\bf Formally:} \\
\smallskip
\quad Modeling of a dependent variable $Y$ by ``learning'' a recursive partition \\
\quad w.r.t explanatory variables $X_1$, \dots, $X_m$.

@@

@@

{\bf Key features:} \\
\begin{itemize}
\item Predictive power in nonlinear regression relationships.
\item Interpretability (enhanced by visualization), i.e., no ``black box'' methods.
\end{itemize}

@@

\begin{frame}[fragile]
\frametitle{Motivation}
{\small Titanic:  \quad $Survived \; \sim  \; Gender + Age + Class$}
\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<ctree_Titanic, echo=FALSE, fig=TRUE, width=11>>=
ct_ttnc <- ctree(Survived ~ Gender + Age + Class, data = ttnc, alpha = 0.01)
plot(ct_ttnc)
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}
{\bf Data:}
The Titanic data set provides information on the survival of 2201~passengers on the Titanic with the additional variables Class, Sex and Age.

@@

The variables and their levels are as follows:

@@

\begin{tabular}{l l l}
\hline 
No & Name &	Levels\\
\hline 
1 &	Class &	1st, 2nd, 3rd, Crew\\
2 &	Sex &	Male, Female\\
3 &	Age &	Child, Adult\\
4 &	Survived &	No, Yes \\
\hline 
\end{tabular}

\end{frame}
\begin{frame}[fragile]
\frametitle{Motivation}
{\small Wages:  \quad $log(wage) \; \sim  \; education + experience + age + ethnicity + gender + union$}
\begin{center}
\setkeys{Gin}{width=0.99\textwidth}
<<ctree_Wages1985, echo=FALSE, fig=TRUE, width=13>>=
ct_cps <- ctree(log(wage) ~ education + experience + age + ethnicity + gender + union, data = CPS1985, alpha = 0.01)
plot(ct_cps)
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}
{\small Wages:  \quad $log(wage) \; \sim  \; education \ | \ experience + age + ethnicity + gender + union$}
\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<lmtree_Wages1985, echo=FALSE, fig=TRUE, width=11>>=
mob_cps <- lmtree(log(wage) ~ education | experience + age + ethnicity + gender + union, data = CPS1985)
plot(mob_cps)
@
\end{center}
\end{frame}

Wind direction forecasting\\
at Innsbruck airport

\vspace{-1.4cm}

\setkeys{Gin}{width=0.95\linewidth}
\begin{center}
\includegraphics{circtree_ibk.pdf}
\end{center}

@@

{\bf Homogeneity:} Partitioning the data into subgroups aims at allowing for a better model fit in each resulting subset.

@@

{\bf Model:} The model fitted to a response $Y$ in each subgroup optimizes an objective function, i.e., it can be a simple constant fit (e.g. average value or proportion, minimizing deviations) but also a more complex model (possibly including additional regressors, and, e.g., minimizing the residual sum of squares or the negative log-likelihood).

@@

{\bf Trade-off:} Complexity in the tree structure vs.\ complexity in the fitted models.\\
Simple models in the nodes (such as constant fits) often lead to large trees while more complex models can compensate a higher variation within the nodes allowing for less complex trees.

% \begin{itemize}
% \item Choose an indicator or a measure of homogeneity within subgroups.
% %% FIXME: discrepancy measure, loss function
% \item Find the split which optimizes this measure in the resulting subgroups.
% \end{itemize}
% 
% \bigskip
% 
% \bigskip
% 
% {\bf Examples of discrepancy measures:} 
% \begin{itemize}
% \item For regression tree (numeric response): Deviations from an average value, residual sum of squares.
% \item For classification tree (categorical response): Number of misclassifications.
% \end{itemize}
 \end{frame}

@@

{\bf Base algorithm:}
\begin{enumerate}
\item Fit a model to the response $Y$.
\item Assess association of $Y$ (or a corresponding transformation / goodness-of-fit measure) and each possible split variable $X_j$.
\item Split sample along the $X_{j^{\ast}}$ with strongest association: Choose breakpoint with highest improvement of the model fit.
\item Repeat steps 1--3 recursively in the subsamples until some stopping criterion is met.
\item Optionally: Reduce the size of the tree through post-pruning, i.e., by cutting off branches that do not improve the performance significantly.
\end{enumerate}

@@

{\bf Examples of possible transformations of the response $Y$:} 
\begin{itemize}
\item Numeric response: 
\begin{itemize}
\item Ranks of $Y$.
\item (Absolute) deviations $Y - \bar Y$.
\item Residual sum of squares $\sum (Y - \hat{Y})^2$.
\end{itemize}
\item Categorical response: 
\begin{itemize}
\item Number of misclassifications (Gini impurity).
\end{itemize}
\end{itemize}

@@

{\bf Possible association tests:} 
$\chi^2$-test, Pearson correlation test, two-sample t-test, ANOVA, signed-rank tests, rank-sum tests, \ldots (depending on the scale and type of variables)

@@

{\bf Split point selection:}
Exhaustive search over all possible split points or two-sample test statistics.

@@

{\bf Stopping criteria:}
Minimal number of observations per node, tree depth, minimal $p$-value, \ldots

@@

{\bf How and where to split?}

@@

\subsection{Split variable selection}
\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Titanic): \ Survived $\sim$ Gender

@@

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Titanic): \ Survived $\sim$ Age

@@

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Titanic): \ Survived $\sim$ Class

@@

% 
% \begin{frame}[fragile]
% \frametitle{Split variable selection} 
% {\bf Compare pairwise associations} (Titanic): 
% Based on independence tests, e.g., $\chi^2$-test for categorical variables.
% \vspace{0.4cm}
% <<echo=TRUE, results=verbatim>>=
% chisq_test(Survived ~ Gender, data = ttnc)
% @
% 
% \vspace{0.2cm}
% 
% <<echo=TRUE, results=verbatim>>=
% chisq_test(Survived ~ Age, data = ttnc)
% @
% 
% \vspace{0.7cm}
% 
% $\Rightarrow$ Select the split variable showing the lowest $p$-value.
% 
% \end{frame}

@@

{\bf Idea:}
Employ statistical tests for the split variable selection.\\
Evaluating associations on the unified $p$-value scale allows for an unbiased comparison of possible split variables on different scales and of different types.

@@

{\bf Unbiased recursive partitioning:} Separation of split variable selection (based on $p$-values) and split point selection (only within the already chosen split variable).

@@

{\bf Alternative:} Directly compare all available split points in all possible split variables. In order to allow for an unbiased selection within this strategy, it has to be accounted for differences such as a varying number of available split points, different scales or variable types by correcting the measurement considered for comparison accordingly. However, this has often not been included in classic tree algorithms.
\end{frame}

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ education

@@

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ gender

@@

% \begin{frame}[fragile]
% \frametitle{Tree algorithm}
% \vspace{0.2cm}
% 
% {\bf How and where to split?}
% 
% \vspace{0.4cm}
% 
% \begin{enumerate}
% %\item \textcolor{gray}{Selection of the best split variable.} \checkmark
% \item Selection of the best split variable.
% \vspace{0.2cm}
% \item Selection of the corresponding best split point.
% \end{enumerate}
% \end{frame}

@@

\begin{frame}[fragile]
\frametitle{Split point selection}
{\bf Exhaustive search:} For each possible split point evaluate a defined objective function in the resulting subgroups and select the split point that minimizes the objective function, i.e., leads to the highest improvement of the model fit.

@@

{\bf Alternative:} Two-sample test statistics evaluating the discrepancy between resulting subgroups. 
\end{frame}

@@

{\bf Goal:} Avoid overfitting.

@@

{\bf Pre-pruning:} Internal stopping criterium. Stop splitting when there is no significant association to any of the possible split variable.

@@

{\bf Post-pruning:} Grow large tree and prune splits that do not improve the model fit (e.g., via cross-validation or information criteria).

@@

{\bf Pre-pruning:}
\begin{itemize}
\item[+] Does not require additional calculations as only already provided information from statistical test is used.
\item[+] Computationally less expensive as trees are prevented from getting too large.
\item[-] Performance depends on the power of the selected statistical test.
\end{itemize}

@@

{\bf Post-pruning:}
\begin{itemize}
\item[+] Stable method to avoid overfitting, regardless of the power of the employed statistical test.
\item[+] Can employ information criteria such as AIC or BIC for model-based partitioning. 
\item[-] Computationally more expensive as very large trees are grown and additional (out-of-bag) evaluations are required.
\end{itemize}

@@

{\bf Note:} The CTree algorithm has been employed for all previous tree illustrations.

@@

{\bf Predictions:} For a (possibly new) set of observations the method \fct{predict} can be employed to obtain the corresponding predicted values of the response variable.

@@

{\bf Post-pruning:} Cost-complexity pruning of an rpart-object with complexity parameter \texttt{cp}.

@@

% \begin{frame}
% \frametitle{R software for tree models}
% 
% {\bf rpart:} 
% \begin{itemize}
% \item R package for Recursive PARTitioning.
% \item Implementation based on ideas of CART (Classification And Regression Trees).
% \item Classification trees: reduce number of misclassifications.
% \item Regression trees: reduce objective function for numeric response.
% \end{itemize}
% 
% \vspace{0.4cm}
% 
% {\bf partykit:} 
% \begin{itemize} 
% \item R package offering a toolkit for unbiased recursive partitioning. 
% \item Implementations of the tree algorithms CTree (Conditional Inference Trees) and MOB (Model-based Recursive Partitioning).
% \item Provides a very general framework with a unifying infrastructure.
% \item Can import rpart.
% \end{itemize}
% 
% \end{frame}

@@

{\bf Further R packages based on partykit:}
\begin{itemize}
  \item evtree (evolutionary trees),
  \item model4you (personalised trees),
  \item disttree (distributional trees),
  \item stablelearner (stability assessment),
  \item \ldots
\end{itemize}
\end{frame}

@@

{\bf Further R packages:}
\begin{itemize}
  \item mvpart (multivariate CART),
  \item RWeka (J4.8, M5', LMT),
  \item and many more (C50, quint, stima, \dots).
\end{itemize}
\end{frame}

>>>>

\textbf{Idea:} ``Divide and conquer.''
\begin{itemize}
  \item \emph{Goal:} Split the data into small(er) and (rather) homogenous subgroups.
  \item \emph{Inputs:} Explanatory variables (or covariates/regressors) used for splitting.
  \item \emph{Output:} Prediction for dependent (or target) variable(s).
\end{itemize}

\bigskip
\pause

\textbf{Formally:}
\begin{itemize}
  \item Dependent variable $Y$ (possibly multivariate).
  \item Based on explanatory variables $X_1, \dots, X_m$.
  \item ``Learn'' subgroups of data by combining splits in $X_1, \dots, X_m$.
  \item Predict $Y$ with (simple) model in the subgroups, often simply the mean.
\end{itemize}

\bigskip
\pause

\textbf{Key features:}
\begin{itemize}
  \item Predictive power in nonlinear regression relationships.
  \item Interpretability (enhanced by tree visualization), i.e., no ``black box''.
\end{itemize}

\textbf{Example:} Survival (yes/no) on the Titanic. Women and children first?

\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<ctree_Titanic, echo=FALSE, fig=TRUE, width=11>>=
ct_ttnc <- ctree(Survived ~ Gender + Age + Class, data = ttnc, alpha = 0.01)
plot(ct_ttnc)
@
\end{center}

\textbf{Model formula:} \code{Survived ~ Gender + Age + Class}.

@@

\textbf{Data:} Information on the survival of 2201~passengers on board the
ill-fated maiden voyage of the RMS~Titanic in 1912. A data frame containing
\Sexpr{nrow(ttnc)} observations on \Sexpr{ncol(ttnc)} variables.

@@

\begin{tabular}{ll}
\hline 
Variable        & Description\\
\hline 
\code{Class}    & Factor: 1st, 2nd, 3rd, or Crew.\\
\code{Gender}   & Factor: Male, Female.\\
\code{Age}      & Factor: Child, Adult.\\
\code{Survived} & Factor: No, Yes.\\
\hline 
\end{tabular}

\textbf{Example:} Determinants of wages and returns to education.

\begin{center}
\setkeys{Gin}{width=0.99\textwidth}
<<ctree_Wages1985, echo=FALSE, fig=TRUE, width=13>>=
ct_cps <- ctree(log(wage) ~ education + experience + age + ethnicity + gender + union, data = CPS1985, alpha = 0.01)
plot(ct_cps)
@
\end{center}

\textbf{Model formula:} \code{log(wage) ~ education + experience + age + ethnicity + gender + union}.

@@

\textbf{Data:} Random sample from the May 1985 US Current Population Survey.
A data frame containing \Sexpr{nrow(CPS1985)} observations on \Sexpr{ncol(CPS1985)} variables.

@@

\begin{tabular}{ll}
\hline 
Variable & Description\\
\hline 
\code{wage}       & Wage (in US dollars per hour). \\
\code{education}  & Education (in years).  \\
\code{experience} & Potential work experience (in years, \code{age - education - 6}). \\
\code{age}        & Age (in years). \\
\code{ethnicity}  & Factor: Caucasian, Hispanic, Other. \\
\code{gender}     & Factor: Male, Female. \\
\code{union}      & Factor. Does the individual work on a union job? \\
\hline 
\end{tabular}

@@

\textbf{Alternatively:} Linear regression tree.

\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<lmtree_Wages1985, echo=FALSE, fig=TRUE, width=11>>=
mob_cps <- lmtree(log(wage) ~ education | experience + age + ethnicity + gender + union, data = CPS1985)
plot(mob_cps)
@
\end{center}

\textbf{Model formula:} \code{log(wage) ~ education | experience + age + ethnicity + gender + union}.

@@

\textbf{Model fit:}
\begin{itemize}
  \item Not just a group-specific mean.
  \item But group-specific intercept and slope (i.e., returns to education).
\end{itemize}

\textbf{Example:} Nowcasting (1--3 hours ahead) of wind direction at Innsbruck Airport.

\setkeys{Gin}{width=0.87\linewidth}
\begin{center}
\includegraphics{circtree_ibk.pdf}
\end{center}

\textbf{Data:} 41,979 data points for various weather observations.
\begin{itemize}
  \item Dependent variable: Wind direction 1--3 hours ahead.
  \item Explanatory variables: Current weather observations including
    wind direction, wind speed, temperature, (reduced) air pressure, relative humidity.
  \item Circular response in $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$.
\end{itemize}

@@

\textbf{Model fit:} Circular distribution (von Mises), fitted by maximum likelihood.

@@

\textbf{Homogeneity:} Partitioning the data into subgroups aims at allowing for a better model fit in each resulting subset.

@@

\textbf{Model:} The model fitted to a response $Y$ in each subgroup optimizes an objective function, i.e., it can be a simple constant fit (e.g. average value or proportion, minimizing deviations) but also a more complex model (possibly including additional regressors, and, e.g., minimizing the residual sum of squares or the negative log-likelihood).

@@

\textbf{Trade-off:} Complexity in the tree structure vs.\ complexity in the fitted models.\\
Simple models in the nodes (such as constant fits) often lead to large trees while more complex models can compensate a higher variation within the nodes allowing for less complex trees.

% \begin{itemize}
% \item Choose an indicator or a measure of homogeneity within subgroups.
% %% FIXME: discrepancy measure, loss function
% \item Find the split which optimizes this measure in the resulting subgroups.
% \end{itemize}
% 
% \bigskip
% 
% \bigskip
% 
% \textbf{Examples of discrepancy measures:} 
% \begin{itemize}
% \item For regression tree (numeric response): Deviations from an average value, residual sum of squares.
% \item For classification tree (categorical response): Number of misclassifications.
% \end{itemize}
 \end{frame}

@@

\textbf{Base algorithm:}
\begin{enumerate}
\item Fit a model to the response $Y$.
\item Assess association of $Y$ (or a corresponding transformation / goodness-of-fit measure) and each possible split variable $X_j$.
\item Split sample along the $X_{j^{\ast}}$ with strongest association: Choose breakpoint with highest improvement of the model fit.
\item Repeat steps 1--3 recursively in the subsamples until some stopping criterion is met.
\item Optionally: Reduce the size of the tree through post-pruning, i.e., by cutting off branches that do not improve the performance significantly.
\end{enumerate}

@@

\textbf{Examples of possible transformations of the response $Y$:} 
\begin{itemize}
\item Numeric response: 
\begin{itemize}
\item Ranks of $Y$.
\item (Absolute) deviations $Y - \bar Y$.
\item Residual sum of squares $\sum (Y - \hat{Y})^2$.
\end{itemize}
\item Categorical response: 
\begin{itemize}
\item Number of misclassifications (Gini impurity).
\end{itemize}
\end{itemize}

@@

\textbf{Possible association tests:} 
$\chi^2$-test, Pearson correlation test, two-sample t-test, ANOVA, signed-rank tests, rank-sum tests, \ldots (depending on the scale and type of variables)

@@

\textbf{Split point selection:}
Exhaustive search over all possible split points or two-sample test statistics.

@@

\textbf{Stopping criteria:}
Minimal number of observations per node, tree depth, minimal $p$-value, \ldots

@@

\textbf{How and where to split?}

@@

\subsection{Split variable selection}
\begin{frame}[fragile]
\frametitle{Split variable selection}
\textbf{Assess pairwise associations} (Titanic): \ Survived $\sim$ Gender

@@

\begin{frame}[fragile]
\frametitle{Split variable selection}
\textbf{Assess pairwise associations} (Titanic): \ Survived $\sim$ Age

@@

\begin{frame}[fragile]
\frametitle{Split variable selection}
\textbf{Assess pairwise associations} (Titanic): \ Survived $\sim$ Class

@@

% 
% \begin{frame}[fragile]
% \frametitle{Split variable selection} 
% \textbf{Compare pairwise associations} (Titanic): 
% Based on independence tests, e.g., $\chi^2$-test for categorical variables.
% \vspace{0.4cm}
% <<echo=TRUE, results=verbatim>>=
% chisq_test(Survived ~ Gender, data = ttnc)
% @
% 
% \vspace{0.2cm}
% 
% <<echo=TRUE, results=verbatim>>=
% chisq_test(Survived ~ Age, data = ttnc)
% @
% 
% \vspace{0.7cm}
% 
% $\Rightarrow$ Select the split variable showing the lowest $p$-value.
% 
% \end{frame}

@@

\textbf{Idea:}
Employ statistical tests for the split variable selection.\\
Evaluating associations on the unified $p$-value scale allows for an unbiased comparison of possible split variables on different scales and of different types.

@@

\textbf{Unbiased recursive partitioning:} Separation of split variable selection (based on $p$-values) and split point selection (only within the already chosen split variable).

@@

\textbf{Alternative:} Directly compare all available split points in all possible split variables. In order to allow for an unbiased selection within this strategy, it has to be accounted for differences such as a varying number of available split points, different scales or variable types by correcting the measurement considered for comparison accordingly. However, this has often not been included in classic tree algorithms.
\end{frame}

\begin{frame}[fragile]
\frametitle{Split variable selection}
\textbf{Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ education

@@

\begin{frame}[fragile]
\frametitle{Split variable selection}
\textbf{Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ gender

@@

% \begin{frame}[fragile]
% \frametitle{Tree algorithm}
% \vspace{0.2cm}
% 
% \textbf{How and where to split?}
% 
% \vspace{0.4cm}
% 
% \begin{enumerate}
% %\item \textcolor{gray}{Selection of the best split variable.} \checkmark
% \item Selection of the best split variable.
% \vspace{0.2cm}
% \item Selection of the corresponding best split point.
% \end{enumerate}
% \end{frame}

@@

\begin{frame}[fragile]
\frametitle{Split point selection}
\textbf{Exhaustive search:} For each possible split point evaluate a defined objective function in the resulting subgroups and select the split point that minimizes the objective function, i.e., leads to the highest improvement of the model fit.

@@

\textbf{Alternative:} Two-sample test statistics evaluating the discrepancy between resulting subgroups. 
\end{frame}

@@

\textbf{Goal:} Avoid overfitting.

@@

\textbf{Pre-pruning:} Internal stopping criterium. Stop splitting when there is no significant association to any of the possible split variable.

@@

\textbf{Post-pruning:} Grow large tree and prune splits that do not improve the model fit (e.g., via cross-validation or information criteria).

@@

\textbf{Pre-pruning:}
\begin{itemize}
\item[+] Does not require additional calculations as only already provided information from statistical test is used.
\item[+] Computationally less expensive as trees are prevented from getting too large.
\item[-] Performance depends on the power of the selected statistical test.
\end{itemize}

@@

\textbf{Post-pruning:}
\begin{itemize}
\item[+] Stable method to avoid overfitting, regardless of the power of the employed statistical test.
\item[+] Can employ information criteria such as AIC or BIC for model-based partitioning. 
\item[-] Computationally more expensive as very large trees are grown and additional (out-of-bag) evaluations are required.
\end{itemize}

@@

\textbf{Note:} The CTree algorithm has been employed for all previous tree illustrations.

@@

\textbf{Predictions:} For a (possibly new) set of observations the method \fct{predict} can be employed to obtain the corresponding predicted values of the response variable.

@@

\textbf{Post-pruning:} Cost-complexity pruning of an rpart-object with complexity parameter \texttt{cp}.

@@

% \begin{frame}
% \frametitle{R software for tree models}
% 
% \textbf{rpart:} 
% \begin{itemize}
% \item R package for Recursive PARTitioning.
% \item Implementation based on ideas of CART (Classification And Regression Trees).
% \item Classification trees: reduce number of misclassifications.
% \item Regression trees: reduce objective function for numeric response.
% \end{itemize}
% 
% \vspace{0.4cm}
% 
% \textbf{partykit:} 
% \begin{itemize} 
% \item R package offering a toolkit for unbiased recursive partitioning. 
% \item Implementations of the tree algorithms CTree (Conditional Inference Trees) and MOB (Model-based Recursive Partitioning).
% \item Provides a very general framework with a unifying infrastructure.
% \item Can import rpart.
% \end{itemize}
% 
% \end{frame}

@@

\textbf{Further R packages based on partykit:}
\begin{itemize}
  \item evtree (evolutionary trees),
  \item model4you (personalised trees),
  \item disttree (distributional trees),
  \item stablelearner (stability assessment),
  \item \ldots
\end{itemize}
\end{frame}

@@

\textbf{Further R packages:}
\begin{itemize}
  \item mvpart (multivariate CART),
  \item RWeka (J4.8, M5', LMT),
  \item and many more (C50, quint, stima, \dots).
\end{itemize}
\end{frame}
